---
title: "Estimating Chronic Disease Disparity, Chicago Metro and CCDPH Jurisdiction"
author: "C. Scott Smith, PhD AICP"
date: "6/14/2021"
output: bookdown:html_document2
tidy: styler
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{r packages_settings, include=FALSE}

library(dplyr)
library(tidyverse)
library(sf)
library(sp)
library(spdep)
library(leaflet)
library(leaflet.providers)
library(kableExtra)
library(Hmisc)
library(bookdown)

# Change the presentation of decimal numbers to 4 and avoid scientific notation
options(digits=4, scipen=999)
```


```{r import_census, include=FALSE}

# import TIGER/Line cartographic boundary shapefiles
INIL_Tracts_geom <- st_read("layers/INIL_Tracts_geom.shp")
INIL_Places_geom <- st_read("layers/INIL_Places_geom.shp")
INIL_Counties_geom <- st_read("layers/INIL_Counties_geom.shp")

# subset counties, rename variables
studyarea_counties <- INIL_Counties_geom %>% 
  filter(STATEFP=="17" &
         GEOID=="17031" | GEOID=="17043" | GEOID=="17089" | GEOID=="17093" | GEOID=="17097" | GEOID=="17111" | GEOID=="17197") %>%
  rename(geoid_county = GEOID,
         state_county = STATEFP,
         name_county = NAME) %>%
  mutate(sqmi_land_county = ALAND*0.00000038610,
         sqmi_water_county = AWATER*0.00000038610,
         sqmi_county = sqmi_land_county+sqmi_water_county) %>%
  select(-c(COUNTYFP:AFFGEOID, LSAD:AWATER))

# create region analysis boundary
studyarea_region <- studyarea_counties %>% 
  st_union() %>% 
  st_as_sf() %>% 
  mutate(studyarea="region")

# subset places using region analysis boundary
studyarea_places <- INIL_Places_geom %>% 
  mutate(area_tot = st_area(geometry)) %>%
  st_intersection(studyarea_region) %>%
  mutate(area_int = st_area(geometry),
         area_int_pct = as.numeric(area_int/area_tot*100)) %>%
  filter(area_int_pct >= 1) %>%
  st_cast("MULTIPOLYGON")

# subset study area tracts data using study area county fips
studyarea_tracts <- INIL_Tracts_geom %>% 
  mutate(geoid_county = paste0(STATEFP,COUNTYFP)) %>%
  rename(geoid_tract = GEOID,
         sqmi_land_tract = sqmi_land,
         sqmi_water_tract = sqmi_water) %>%
  left_join(studyarea_counties %>% st_drop_geometry(), by="geoid_county") %>%
  drop_na(name_county) %>%
  distinct(geoid_tract,.keep_all = TRUE)

# remove raw data
rm(INIL_Places_geom, INIL_Counties_geom, INIL_Tracts_geom) 
```

```{r import_places, include=FALSE}

# import PLACES chronic disease data
places_tracts <- read_csv("data/places_tracts.csv")
places_datadictionary <- read_csv("data/places_datadictionary.csv") %>%
  select(category:long_desc) %>%
  mutate(variable = tolower(str_replace_all(name, "_CrudePrev", "")),
         variable = tolower(str_replace_all(variable, "_crude95ci", "")))
  
# join study area census tracts with places data
studyarea_tracts_disease <- places_tracts %>% 
  rename(geoid_tract=TractFIPS) %>%
  right_join(studyarea_tracts, by="geoid_tract") %>%
  select(geoid_tract,
         ends_with("CrudePrev"),
         sqmi_land_tract,
         sqmi_water_tract,
         sqmi,
         geometry,
         -c(TotalPopulation,StateAbbr:CountyFIPS)) %>%
  st_as_sf()

rm(places_tracts)
```


```{r import_acs, include = FALSE}

# import race/ethnic data
acs2019_bytract <- read_csv("data/acs2019_bytract.csv") %>%
  mutate(GEOID = as.character(GEOID))

studyarea_counties_acs <- acs2019_bytract %>% 
  mutate(geoid_county = substr(GEOID,1,5)) %>% 
  right_join(studyarea_counties, by="geoid_county") %>%
  group_by(geoid_county) %>%
  summarize_at(.vars = vars(TotalPop:PopWht, TotalPov:PopBPov), .funs=c(sum="sum")) %>%
  left_join(studyarea_counties, by="geoid_county") %>%
  st_as_sf()
```

# Background
This report maps the geographic patterning of chronic health conditions to identify disease hot spots and evaluate differences in outcomes by sociodemographic group within the greater Chicago region. The report finds: (1) that the spatial arrangement and magnitude of geographic clusters vary by category of disease; and (2) considerable disparities in health outcomes exist between racial, ethnic and economic groups both at the regional and subregional scales.

# Study areas
We examine patterns of chronic disease prevalence within two study areas; one multi-county and regional and the other confined to CCDPH's jurisdiction within Cook County. The regional, multiple county analysis examines broader public health patterns across seven counties, specifically, Cook, DuPage, Kane, Kendall, Lake, McHenry and Will county all located in the Chicago metropolitan area of northeastern Illinois. Collectively, this contiguous area spans `r prettyNum(set_units(st_area(studyarea_region$x),mi^2), big.mark = ",")` square miles (`prettyNum(set_units(st_area(studyarea_region$x),km^2), big.mark = ",")` square kilometers) and has a combined residential population of `r prettyNum(sum(studyarea_counties_acs$TotalPop_sum)/1000000, digits=2)` million (ACS, 2019). 

These seven counties were selected in part because they represent considerable diversity both in terms of urban design (e.g., including both higher-density urban centers as well as lower-density exurban communities) and sociodemographic makeup. The counties also correspond with the Chicago Metropolitan Agency for Planning's (CMAP) planning area thereby allowing for ready integration of the metropolitan planning organization's rich datasets and presenting opportunities for future research to be carried across CMAP's jurisdiction. In addition to the regional analysis, we carry out a similar evaluation of disease and disparity within CCDPH's jurisdiction which is wholly contained within Cook County.

```{r study_area_map, fig.cap="Chicago Metropolitan Region Study Area Boundary, Counties"}

labels <- sprintf(
  "<strong>%s</strong><br/>%s mi<sup>2</sup><br/>%s",
  paste0(studyarea_counties_acs$name_county," County"),
  paste0("Area: ", prettyNum(studyarea_counties_acs$sqmi_county,digits=2, big.mark = ",")),
  paste0("Population: ", prettyNum(studyarea_counties_acs$TotalPop_sum/1000000, digits=2), " million")) %>% lapply(htmltools::HTML)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(group = "places",
              data = studyarea_places %>% st_transform(crs=4326),
              fill="grey",
              weight = 0.5,
              color = "white",
              fillOpacity = 0.25) %>%
  addPolygons(group = "county",
              data=studyarea_counties_acs %>% st_transform(crs=4326),
              fillColor = "orange",
              weight = 2,
              opacity = 0.5,
              color = "white",
              dashArray = "3",
              fillOpacity = 0.7,
              highlight = highlightOptions(
                weight = 5,
                color = "#666",
                dashArray = "",
                fillOpacity = 0.7,
                bringToFront = FALSE),
              label = labels,
              labelOptions = labelOptions(
                style = list("font-weight" = "normal", padding = "3px 8px"),
                textsize = "15px",
                direction = "auto")) %>%
      addPolylines(group = "boundary",
              data = studyarea_region %>% st_transform(crs=4326),
              fill=NULL,
              weight = 3,
              color = "black") %>%
      addLayersControl(
          overlayGroups =c("county", "places", "boundary"),
          options = layersControlOptions(collapsed=FALSE)
          )

```

# Data & methodology
We employ three publicly available datasets to carry out the study.

## CDC's PLACES modeled chronic disease prevalence (link)
Centers for Disease Control and Prevention (CDC) PLACES dataset was used to model chronic disease across the study areas at the census tract level. The PLACES Project is an expansion of the original 500 Cities Project that began in 2015. The original project was launched by the Centers for Disease Control and Prevention (CDC) in partnership with the Robert Wood Johnson Foundation (RWJF) and CDC Foundation. In 2018, this partnership was extended through 2020. In 2020, the project expanded to provide small area estimates (SAE) for counties, places, census tracts, and ZIP Code Tabulation Areas (ZCTA) across the United States. The CDC uses a multilevel regression and poststratification (MRP) approach which links geocoded health surveys--specifically, the Behavioral Risk Factor Surveillance System (BRFSS)--with demographic and socioeconomic data made available in the American Community Survey (ACS) to estimate geographic distributions of population disease burden and health behaviors at the community level.

## ACS 5-year sociodemographic estimates (link)
The analysis uses ACS 5-year estimates of study area populations by age, race, ethnicity and income for standardizing prevalence measures across geographies and identifying health disparities by demographic group. The 5-year estimates used in this study represent data collected between 2015 and 2019, which allows for greater  statistical reliability.

## TIGER/Line cartographic boundary data
Topologically Integrated Geographic Encoding and Referencing (TIGER) Database (MTDB) or TIGER/Line shapefiles of the study area were used to link geographies to data in both the ACS and PLACES at the census tract level. The geometries made it possible to estimate global and spatial autocorrelation across sociodemographic and chronic disease variables.

```{r spatial_autocorrelation}

# Create queen neighborhood structure 
tracts_nb <- poly2nb(studyarea_tracts_disease, queen=TRUE)
# tracts_nb

# Define neighborhood weights matrix
tracts_wm <- nb2listw(tracts_nb)
# tracts_wm

health_types <- as.data.frame(colnames(studyarea_tracts_disease)) %>%
  rename(type = 1) %>%
  filter(str_detect(type,"CrudePrev"))

moran_stats <- data.frame(matrix(NA, ncol = 6, nrow = 0)) %>%
  rename(type = 1,
         morani = 2,
         morane = 3,
         moranv = 4,
         moransd = 5,
         moranp = 6) %>%
  mutate(type = as.character(type),
         morani = as.numeric(morani),
         morane = as.numeric(morane),
         moranv = as.numeric(moranv),
         moransd  = as.numeric(moransd),
         moranp = as.numeric(moranp)
         )

for (i in 1:nrow(health_types)) {
  
  studyarea_tracts_type <- studyarea_tracts_disease %>%
    select(health_types$type[i])
  tracts_moran <- moran.test(studyarea_tracts_type[[1]], tracts_wm, na.action = na.omit)
  moran_stats <- moran_stats %>%
    bind_rows(as.data.frame(list(
      health_types$type[i],
      tracts_moran$estimate[[1]],
      tracts_moran$estimate[[2]],
      tracts_moran$estimate[[3]],
      tracts_moran$statistic[[1]],
      tracts_moran$p.value)) %>%
  rename(type = 1,
         morani = 2,
         morane = 3,
         moranv = 4,
         moransd = 5,
         moranp = 6))
}

moran_stats %>% 
  arrange(-morani) %>% 
  select(Name = type,
         "Moran's I" = morani,
         "Standard Deviation" = moransd,
         -morane, 
         -moranv) %>%
    filter(str_detect(Name,"CrudePrev")) %>%
  mutate(Name = tolower(str_replace_all(Name, "_CrudePrev", ""))) %>%
  kbl(caption=paste0("Moran's I Statistics for Chicago Metropolitan Area by Chronic Health Prevalance Category")) %>%
  kable_classic(full_width=F)

```

```{r prevalence_index, include=FALSE}

# create function for standardizing prevalence variables
normalize <- function(x) {
return ((x - min(x)) / (max(x) - min(x)))
}

places_tracts_sa_vars <- studyarea_tracts_disease %>%
# filter(StudyArea == "Outside CCDPH") %>%
  drop_na(ACCESS2_CrudePrev) %>%
  select(geometry,
         geoid_tract,
         sqmi,
         arth_cp = ARTHRITIS_CrudePrev,
         bph_cp = BPHIGH_CrudePrev,
         diab_cp = DIABETES_CrudePrev,
         obes_cp = OBESITY_CrudePrev,
         lpa_cp = LPA_CrudePrev) %>%
mutate(arth_std = normalize(arth_cp),
       bph_std = normalize(bph_cp),
       diab_std = normalize(diab_cp),
       obes_std = normalize(obes_cp),
       lpa_std = normalize(lpa_cp),
       cpi = (arth_std+bph_std+diab_std+obes_std+lpa_std)/5,
       cpi_ntile = ntile(cpi,5),
       countyfp = substr(geoid_tract,1,5)) %>%
  left_join(acs2019_bytract,by=c("geoid_tract"="GEOID")) %>%
  st_as_sf()

st_write(places_tracts_sa_vars, "layers/places_tracts_sa_vars.shp", append=FALSE)
# st_write(places_tracts_sa_vars, "../layers/chronic_conditions_tract_PLACES.shp", append=FALSE)

```


```{r lisa_clusters}

# Generate local moran statistics for all health categories
# Ii: local moran statistic; E.Ii: expectation of local moran statistic; Var.Ii: variance of local moran statistic; Z.Ii: standard deviate of local moran statistic; Pr(): p-value of local moran statistic

local_df_all <- read_rds("data/local_df.rds")

for (i in 1:nrow(health_types)) {
  
  places_tracts_sa_sf_type <- studyarea_tracts_disease %>%
    st_drop_geometry() %>% 
    select(health_types$type[i],
           geoid_tract) %>%
    replace(is.na(.), 0)
  df_test <- places_tracts_sa_sf_type[[1]]
  
  local <- localmoran(x = df_test, listw = nb2listw(tracts_nb, style = "W"))
  
  local_df <- as.data.frame(local) %>%
    bind_cols(places_tracts_sa_sf_type) %>%
    rename(CrudePrev = 6) %>%
    mutate(type = health_types$type[i],
           CrudePrev_std = CrudePrev - mean(CrudePrev),
           Ii_std = Ii - mean(Ii),
           quadrant = case_when(CrudePrev_std>0 & Ii_std >0 ~ 1,
                                CrudePrev_std<0 & Ii_std <0 ~ 3,
                                CrudePrev_std<0 & Ii_std >0 ~ 2,
                                CrudePrev_std>0 & Ii_std <0 ~ 4),
           label = case_when(CrudePrev_std>0 & Ii_std >0 ~ "high-high",
                                CrudePrev_std<0 & Ii_std <0 ~ "low-low",
                                CrudePrev_std<0 & Ii_std >0 ~ "low-high",
                                CrudePrev_std>0 & Ii_std <0 ~ "high-low"),
           quadrant = if_else(`Pr(z > 0)`>0.01,0,quadrant),
           label = if_else(`Pr(z > 0)`>0.01,"not significant",label))
  
  local_df_all <- local_df_all %>%
    bind_rows(local_df)
}

# empty dataframe template
local_df <- local_df[0,]
write_rds(local_df, "data/local_df.rds")

local_sf_all <- local_df_all %>%
  left_join(studyarea_tracts, by="geoid_tract") %>%
  filter(type=="LPA_CrudePrev",
         `Pr(z > 0)`<=0.01) %>% 
  st_as_sf() %>%
  st_transform(crs=4326)

pal_lisa <- colorFactor(
  palette = c('red', '#fadbd8', '#aed6f1', 'blue'),
  local_sf_all$label
)

leaflet() %>% 
  addPolygons(data=local_sf_all,
              fillColor = ~pal_lisa(label),
              weight = 0.1,
              opacity = 0.5,
              color = "white",
              dashArray = "3",
              fillOpacity = 0.7) %>%
    addPolygons(data=studyarea_counties %>% st_transform(crs=4326),
              fill = FALSE,
              weight = 1,
              opacity = 0.5,
              color = "gray",
              dashArray = "3",
              fillOpacity = 0.7) %>%
  addLegend("bottomright", 
            pal = pal_lisa, 
            values=c("high-high","high-low","low-high","low-low"), 
            opacity = 1, 
            title="LISA clusters") %>%
  addProviderTiles(providers$CartoDB.Positron)


```


```{r summary_tables}

places_tracts_sa_vars_in %>% st_drop_geometry() %>% group_by(cpi_ntile) %>% summarise(n())

places_tracts_sa_vars_sum <- places_tracts_sa_vars %>%
  st_drop_geometry() %>%
  mutate(cpi_group = case_when(cpi_ntile==1 ~ "Bottom Quintile",
                               cpi_ntile>=2 & cpi_ntile<=4 ~ "Middle Quintiles",
                               cpi_ntile==5 ~ "Top Quintile")) %>%
  group_by(cpi_group) %>%
  select(TotalPop,
         Pop18Pl,
         cpi,
         sqmi,
         PopWht,
         PopBlk,
         PopLat,
         PopBPov,
         TotalPov,
         arth_cp,
         bph_cp,
         diab_cp,
         obes_cp,
         lpa_cp) %>%
  summarise(NumTracts = n(),
            Population = sum(TotalPop),
            Area = sum(sqmi),
            PopDensity = sum(TotalPop/sqmi*TotalPop/sum(TotalPop)),
            WhiteNL = sum(PopWht)/sum(TotalPop)*100,
            BlackNL = sum(PopBlk)/sum(TotalPop)*100,
            Latinx = sum(PopLat)/sum(TotalPop)*100,
            PopBPov_p = sum(PopBPov)/sum(TotalPov)*100,
            CPI_wt = sum(cpi*Pop18Pl/sum(Pop18Pl))*100,
            Obes_wt = sum(obes_cp*Pop18Pl/sum(Pop18Pl)),
            Diab_wt = sum(diab_cp*Pop18Pl/sum(Pop18Pl)),
            LPA_wt = sum(lpa_cp*Pop18Pl/sum(Pop18Pl)),
            BPH_wt = sum(bph_cp*Pop18Pl/sum(Pop18Pl)),
            Arth_wt = sum(arth_cp*Pop18Pl/sum(Pop18Pl))) %>%
  t()

# Reformat summary table

places_tracts_sa_vars_sum_alt <- cbind(rownames(places_tracts_sa_vars_sum), data.frame(places_tracts_sa_vars_sum, row.names=NULL))

colnames(places_tracts_sa_vars_sum_alt) <- places_tracts_sa_vars_sum_alt[c(1),]
places_tracts_sa_vars_sum_alt <- places_tracts_sa_vars_sum_alt[-c(1),]


places_tracts_sa_vars_sum_alt %>%
  rename(Category = cpi_group) %>%
  mutate_at(vars(`Bottom Quintile`:`Top Quintile`),as.numeric) %>%
  mutate_if(is.numeric, prettyNum, digits=2, big.mark=",") %>%
  kbl(caption=paste0("Chronic Health Prevalance Variable Names, Categories and Descriptions")) %>%
  kable_classic(full_width=F)

```

``` {r correlation_matrix}
variables_corr <- ccdph_tracts_vars %>% 
  st_drop_geometry() %>%
  filter(dist != "outside") %>%
  select(`% below poverty` = PctP_19,
        `hypertension` = bph_cp,
        `diabetes` = diab_cp,
        `obesity` = obes_cp,
        `low physical activity` = lpa_cp,
        `% black`=PpBlk_p,
        `% people of color` = PpClr_p,
        `% latinx` = PpLt_pc,
        `% white` = PpWht_p,
        `% asian` = PpAsn_p) %>%
  drop_na()

M <- cor(variables_corr)

# corrplot(M, type="upper",col=c("black","white"), order="hclust", bg="lightgrey", tl.col = "grey")
corrplot(M, method="number", col=c("black","white"), order="hclust", bg="lightgrey", tl.col = "grey")
corrplot(M, col=c("black","white"), order="hclust", bg="lightgrey", tl.col = "grey")
st_write(cc_ccdph_tracts_pov_chronic, "layers/cc_ccdph_tracts_pov_chronic.shp")

```

```{r appendix}
places_datadictionary %>%
  filter(str_detect(name,"CrudePrev")) %>%
  mutate(name = tolower(str_replace_all(name, "_CrudePrev", ""))) %>%
  select(Category = category, 
         Name = short_desc,
         Variable = variable,
         Description = long_desc) %>%
  arrange(Category) %>%
  kbl(caption=paste0("Chronic Health Prevalance Variable Names, Categories and Descriptions")) %>%
  kable_classic(full_width=F)
```

